Pathogen Spillover - dissertation chapter

### 2018-10-08 pathogen-spillover-diss-ch.Rmd
Revise formatting and output model plots.

## Results

@@

### Path model 
This is the list of models that make up the path model to be fit using the methods implemented in the 'piecewiseSEM' packag. The continuous variables in the models were standardized using the default settings of the 'scale' function in R, resulting in each having a mean of zero and standard deviation of one. Each variable is first centered by subtracting its mean and then scaled by dividing by its standard deviation.

@@

Add some re-calcualted variables and explore the bay density~diversity relationship.
```{r re-examine data}

# refit model with missing path from diversity to umca density
oaks.data <- oakplots.model.data %>% 
      rename(hrs25 = hrs_abv25ds, 
             hrs1422 = avg_hrs1422_wet_t_t1.log, 
             lfct = tot_lfct.log, 
             umcadens = umca_density_ha.log, 
             h2005 = H.2005, 
             oakinf = oak.inf, 
             year = sample_year)

oaks.data <- oaks.data %>% 
      mutate(tot_bay.ln = log(tot_bay),
             tot_bay.ln1 = log1p(tot_bay),
             tot_bay0.1 = ifelse(tot_bay==0, 0.1, tot_bay),
             tot_bay.ln.1 = log(tot_bay0.1)
             )

GGally::ggpairs(select(oaks.data, h2005, twi15m, tot_bay, tot_bay.ln, tot_bay.ln1))

@@

ggplot(oaks.data) +
      geom_density(aes(h2005)) +
      theme_classic()

@@

dat <- filter(oaks.data, year==2006)
m6 <- glm(tot_bay ~ h2005 + twi15m, data = dat, family = quasipoisson, na.action = na.fail)
summary(m6)
# plot(m6)
(c_hat <- deviance(m6) / df.residual(m6))
library(MuMIn)
dredge(m6, rank = "QAIC", chat = c_hat)
dredge(m6, rank = "AIC")
hacked.quasipoisson <- function(...) {
    res <- quasipoisson(...)
    res$aic <- poisson(...)$aic
    res
}
QAIC(update(m6, family = hacked.quasipoisson, chat = c_hat)

@@

# GGally::ggpairs(select(oaks.data, tot_lfct,tot_bay,h2005))
m1 <- glmer(tot_lfct ~ tot_bay + hrs25 + hrs1422 + h2005 + (1|plotid) + (1|year), family = poisson, data = oaks.data, na.action = na.omit)
summary(m1)
m2 <- glmer.nb(tot_lfct ~ tot_bay + hrs25 + hrs1422 + h2005 + (1|plotid) + (1|year), data = oaks.data, na.action = na.omit)
summary(m2)
anova(m2,lfctmod)
plot(lfctmod)
```

@@

### Examining the residual plots of each piece of the path model

**Residual plots of Shannon's Diversity Index model**
```{r shannon.mod plot, echo=F}

library(sjPlot)
plot_model(shannon.mod, type = "diag", show.values = T, grid = TRUE)
plot_model(shannon.mod, type = "slope", show.values = T)
plot_model(shannon.mod, type = "resid", show.values = T)
plot_model(shannon.mod, type = "std", show.values = T, 
           title = "Standardized coefficient estimate")
S.plot <- plot_model(shannon.mod, type = "slope")

@@

**Residual plots of bay laurel density model**
```{r umca.mod plot, echo=F}

plot_model(umca.mod, type = "diag")
plot_model(umca.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - UMCA density")
plot_model(umca.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - UMCA density")
plot_model(umca.mod, show.values = T, type = "re")

@@

**Residual plots of symptomatic leaf count model**
```{r lfct.mod plot, echo=F}

@@

# estimate plots
plot_model(lfct.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - leaf count")
plot_model(lfct.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - leaf count")
plot_model(lfct.mod, show.values = T, type = "re")

@@

**Residual plots of average number of wet-hours model**
```{r wet.mod plot, echo=F}
# diagnostic plots
plot_model(wet.mod, type = "diag")
plot_model(wet.mod, type = "resid")

# estimate plots
plot_model(wet.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - wet hours")
plot_model(wet.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - wet hours")
plot_model(wet.mod, show.values = T, type = "re")

@@

**Residual plots of dry-season temperature >25C model**
```{r hot.mod plot, echo=F}

@@

# estimate plots
plot_model(hot.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - hours >25 ºC")
plot_model(hot.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - hours >25 ºC")
plot_model(hot.mod, show.values = T, type = "re")

# effect plots
plot_model(hot.mod, type = "slope") +
      # geom_point() +
      theme_classic() +
      ylab("Number of hours >25 ºC") +
      xlab("Topographic index") +
      theme(strip.background = element_blank(),
            strip.text = element_blank()) +
      NULL

@@

**Plots of oak infection model**
```{r inf.mod plot, echo=F}

@@

cowplot::plot_grid(
      plot_model(inf.mod, type = "est", show.values = T, 
                 title = "Oak infection - est."),
      plot_model(inf.mod, type = "std", show.values = T, 
                 title = "Oak infection std. est."),
      nrow = 2
      )

plot_model(inf.mod, type = "pred", terms = c("lfct")) +
      # geom_point() +
      theme_classic() +
      xlab("log( symptomatic leaf count + 1 )") +
      ylab("Predicted probability for oak infection") +
      ggtitle("") +
      NULL

ggplot(oaks.data, aes(lfct, inf_oak_ct/(inf_oak_ct+uninf_oak_ct))) +
      geom_point() +
      geom_smooth(method = "loess", color = "black") +
      theme_classic() +
      NULL

@@

## Check for spatial autocorrelation
Check oak infection and symptomatic leaf count response variables and model residuals.

**Oak infection autocorrelation**
```{r spatial autocorrelation inf.mod}

@@

infmod.resids <- resid(infmod)
Moran.I(infmod.resids, pl.dists.inv)
cgram.inf.resids <- correlog(oaks.data$x, oaks.data$y, infmod.resids, increment = 300, resamp = 99)
plot(cgram.inf.resids, main = "Oak infection residuals correlogram")

vario.inf.resids <- variog(coords = oakplots.coords,
                            data = infmod.resids)
plot(vario.inf.resids)
fit.vario.inf.resids <- variofit(vario.inf.resids)
lines(fit.vario.inf.resids)
abline(h = fit.vario.inf.resids$nugget, col = "purple")
abline(h = fit.vario.inf.resids$cov.pars[1], col = "darkblue")
abline( v = -log(.05)*fit.vario.inf.resids$cov.pars[2], col = "tan4")
```

@@

**Symptomatic leaf count**
```{r spatial autocorrelation lfct.mod}
# lfct.data <- oakplots.model.data %>% 
#       select(sample_year, tot_lfct.log) %>% 
#       arrange(sample_year)
# acf(lfct.data$tot_lfct.log)

@@

Moran.I(oaks.data$lfct, pl.dists.inv)
cgram.lfct <- correlog(oaks.data$x, oaks.data$y, oaks.data$tot_lfct.log, increment = 300, resamp = 99)
plot(cgram.lfct, main = "Leaf count response correlogram")

r.lfctmod <- resid(lfctmod)
Moran.I(r.lfctmod, pl.dists.inv)
cgram.lfct.resids <- correlog(oaks.data$x, oaks.data$y, r.lfctmod, increment = 300, resamp = 99)
plot(cgram.lfct.resids, main = "Leaf count residuals correlogram")

@@

### Fitting the path model

@@

### Path coefficients
The path model provides direct and indirect effects of predictor variables on multiple responses, which can be combined to calculate the total effect of a predictor variable on a response. Direct effects are from the variables that have a path pointing directly to a response variable, so effect on the response is the value of the path coefficient. Variables with indirect effects are those that are at least once removed from another response variable, that is the response variable of a predictor with an indirect effect is the predictor variable along another path in the model. Indirect effects are calculated by multiplying the coefficients along the path to the response variable of interest. The total effect of variables that have a direct and indirect effect(s) on a response is the sum of the direct and indirect effects. The standardized coefficients of the predictors that have the same response variable (i.e. direct effects) can be interpreted relative to each other in terms of influence on the response variable.

@@

```{r plot rainfall figure, echo=F}
library(ggplot2)
rainfall
# summary(rainy_season_ppt)
# rainy_season_ppt$year <- as.factor(rainy_season_ppt$year)
rainfall_figure <- ggplot(filter(rainy_season_ppt, year > 2004), 
                          aes(year, rain2d_total))
rainfall_figure +
      geom_boxplot(aes(group = year), notch = T) +
      geom_hline(aes(yintercept = mean(rain2d_total)), linetype = "dashed") +
      theme_bw() +
      xlab("Sample Year") +
      ylab("Total Rainfall (mm)") +
      # ggtitle("Rainfall Estimated at Plots from Rain Gauges (November - May)") +
      theme(panel.grid = element_blank(),
            axis.title = element_text(size = 12),
            axis.text = element_text(size = 10))
# ggsave("figures/rainy-season-ppt.eps", width = 6.25, height = 4.5)

>>>>

@@

# Path modeling 
This is the list of models that make up the path model to be fit using the methods implemented in the 'piecewiseSEM' packag. The continuous variables in the models were standardized using the default settings of the 'scale' function in R, resulting in each having a mean of zero and standard deviation of one. Each variable is first centered by subtracting its mean and then scaled by dividing by its standard deviation.

@@

Add some re-calcualted variables and explore the bay density~diversity relationship.

```{r revise data names}
oaks.data <- oakplots.model.data %>% 
      rename(hrs25 = hrs_abv25ds, 
             hrs1422 = avg_hrs1422_wet_t_t1.log, 
             lfct = tot_lfct.log, 
             umcadens = umca_density_ha.log, 
             h2005 = H.2005, 
             oakinf = oak.inf, 
             year = sample_year)

oaks.data <- oaks.data %>% 
      mutate(tot_bay.ln = log(tot_bay),
             tot_bay.ln1 = log1p(tot_bay),
             tot_bay0.1 = ifelse(tot_bay==0, 0.1, tot_bay),
             tot_bay.ln.1 = log(tot_bay0.1)
             )
```

```{r explore diveristy density}

## refit model with missing path from diversity to umca density

library(GGally)
ggpairs(select(oaks.data, h2005, twi15m, tot_bay, tot_bay.ln, tot_bay.ln1))

@@

ggplot(oaks.data) +
      geom_density(aes(h2005)) +
      theme_classic()
```

```{r explore models, eval=F}

@@

dat <- filter(oaks.data, year==2006)
m6 <- glm(tot_bay ~ h2005 + twi15m, data = dat, family = quasipoisson, na.action = na.fail)
summary(m6)
# plot(m6)
(c_hat <- deviance(m6) / df.residual(m6))

library(MuMIn)
dredge(m6, rank = "QAIC", chat = c_hat)
dredge(m6, rank = "AIC")

hacked.quasipoisson <- function(...) {
    res <- quasipoisson(...)
    res$aic <- poisson(...)$aic
    res
}

QAIC(update(m6, family = hacked.quasipoisson, chat = c_hat)

@@

# GGally::ggpairs(select(oaks.data, tot_lfct,tot_bay,h2005))

m1 <- glmer(tot_lfct ~ tot_bay + hrs25 + hrs1422 + h2005 + (1|plotid) + (1|year), family = poisson, data = oaks.data, na.action = na.omit)
summary(m1)

m2 <- glmer.nb(tot_lfct ~ tot_bay + hrs25 + hrs1422 + h2005 + (1|plotid) + (1|year), data = oaks.data, na.action = na.omit)

summary(m2)
anova(m2,lfctmod)
plot(lfctmod)

@@

# Examining the residual plots of each piece of the path model

## **Residual plots of Shannon's Diversity Index model**

### Diagnostic plots 
```{r shannon.mod diagnostic plots, echo=F}

library(sjPlot)
plot_model(shannon.mod, type = "diag", show.values = T, grid = TRUE)
plot_model(shannon.mod, type = "resid", show.values = T)

### Fitted model plots
```{r shannon.mod fit plots}

plot_model(shannon.mod, type = "slope", show.values = T)
plot_model(shannon.mod, type = "std", show.values = T, 
           title = "Standardized coefficient estimate")
S.plot <- plot_model(shannon.mod, type = "slope")

@@

## **Residual plots of bay laurel density model**

### Diagnostic plots
```{r umca.mod diagnostic plots, echo=F}

plot_model(umca.mod, type = "diag")
plot_model(umca.mod, type = "resid")

### Fitted model plots
```{r umca.mod fit plots}

plot_model(umca.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - UMCA density")
plot_model(umca.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - UMCA density")
plot_model(umca.mod, show.values = T, type = "re")

@@

## **Residual plots of symptomatic leaf count model**

### Diagnostic plots
```{r lfct.mod plot, echo=F}

@@

### Fitted model plots
```{r lfct.mod fit plots}
# estimate plots
plot_model(lfct.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - leaf count")
plot_model(lfct.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - leaf count")
plot_model(lfct.mod, show.values = T, type = "re")

@@

## **Residual plots of average number of wet-hours model**

### Diagnostic plots
```{r wet.mod diganostic plots, echo=F}
# diagnostic plots
plot_model(wet.mod, type = "diag")
plot_model(wet.mod, type = "resid")

### Fitted model plots
```{r wet.mod fit plots}
# estimate plots
plot_model(wet.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - wet hours")
plot_model(wet.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - wet hours")
plot_model(wet.mod, show.values = T, type = "re")

@@

## **Residual plots of dry-season temperature >25C model**

### Diagnostic plots
```{r hot.mod diagnostic plots, echo=F}

@@

### Fitted model plots
```{r hot.mod fit plots}
# estimate plots
plot_model(hot.mod, show.values = T, type = "est", 
           title = "Coefficient estimates - hours >25 ºC")
plot_model(hot.mod, show.values = T, type = "std", 
           title = "Standardized coefficient estimates - hours >25 ºC")
plot_model(hot.mod, show.values = T, type = "re")

# effect plots
plot_model(hot.mod, type = "slope", show.data = T) +
      # geom_point() +
      theme_classic() +
      ylab("Number of hours >25 ºC") +
      xlab("Topographic index") +
      theme(strip.background = element_blank(),
            strip.text = element_blank()) +
      NULL

@@

## **Plots of oak infection model**

### Diagnostic plots
```{r inf.mod diagnostic plots, echo=F}

@@

### Fitted model plots
```{r inf.mod fit plots, echo=F}

cowplot::plot_grid(
      plot_model(inf.mod, type = "est", show.values = T, 
                 title = "Oak infection coefficient estimates") +
            # geom_hline(yintercept = 1, color = "gray80",linetype = "dashed") +
            theme_classic(),
      plot_model(inf.mod, type = "std", show.values = T, 
                 title = "Oak infection std. coefficient estimates",
                 # vline.color = "gray80"
                 ) +
            theme_classic(),
      nrow = 2
      )

plot_model(inf.mod, type = "pred", terms = c("lfct"), show.data = T) +
      # geom_point() +
      theme_classic() +
      xlab("log( symptomatic leaf count + 1 )") +
      ylab("Predicted probability for oak infection") +
      ggtitle("") +
      NULL

ggplot(oaks.data, aes(exp(lfct), inf_oak_ct/(inf_oak_ct+uninf_oak_ct))) +
      # geom_point(alpha = .5) +
      stat_sum() +
      facet_grid(year~.) +
      geom_smooth(method = "loess", color = "black") +
      theme_classic() +
      NULL

@@

# Check for spatial autocorrelation
Check oak infection and symptomatic leaf count response variables and model residuals.

## **Oak infection autocorrelation**
```{r spatial autocorrelation inf.mod}

@@

inf.mod.resids <- resid(inf.mod)
Moran.I(inf.mod.resids, pl.dists.inv)
cgram.inf.resids <- correlog(oaks.data$x, oaks.data$y, inf.mod.resids, increment = 300, resamp = 99)
plot(cgram.inf.resids, main = "Oak infection residuals correlogram")

vario.inf.resids <- variog(coords = oakplots.coords,
                            data = inf.mod.resids)
plot(vario.inf.resids)
fit.vario.inf.resids <- variofit(vario.inf.resids)
lines(fit.vario.inf.resids)
abline(h = fit.vario.inf.resids$nugget, col = "purple")
abline(h = fit.vario.inf.resids$cov.pars[1], col = "darkblue")
abline( v = -log(.05)*fit.vario.inf.resids$cov.pars[2], col = "tan4")
```

@@

## **Symptomatic leaf count**
```{r spatial autocorrelation lfct.mod}

# lfct.data <- oakplots.model.data %>% 
#       select(sample_year, tot_lfct.log) %>% 
#       arrange(sample_year)
# acf(lfct.data$tot_lfct.log)

@@

Moran.I(oaks.data$lfct, pl.dists.inv)
cgram.lfct <- correlog(oaks.data$x, oaks.data$y, oaks.data$lfct, increment = 300, resamp = 99)
plot(cgram.lfct, main = "Leaf count response correlogram")

r.lfctmod <- resid(lfct.mod)
Moran.I(r.lfctmod, pl.dists.inv)
cgram.lfct.resids <- correlog(oaks.data$x, oaks.data$y, r.lfctmod, increment = 300, resamp = 99)
plot(cgram.lfct.resids, main = "Leaf count residuals correlogram")

@@

# Fitting the path model

@@

## Path coefficients
The path model provides direct and indirect effects of predictor variables on multiple responses, which can be combined to calculate the total effect of a predictor variable on a response. Direct effects are from the variables that have a path pointing directly to a response variable, so effect on the response is the value of the path coefficient. Variables with indirect effects are those that are at least once removed from another response variable, that is the response variable of a predictor with an indirect effect is the predictor variable along another path in the model. Indirect effects are calculated by multiplying the coefficients along the path to the response variable of interest. The total effect of variables that have a direct and indirect effect(s) on a response is the sum of the direct and indirect effects. The standardized coefficients of the predictors that have the same response variable (i.e. direct effects) can be interpreted relative to each other in terms of influence on the response variable.

@@

```{r plot rainfall figure, echo=F}
library(ggplot2)
# rainfall
# summary(rainy_season_ppt)
# rainy_season_ppt$year <- as.factor(rainy_season_ppt$year)
rainfall_figure <- ggplot(filter(rainy_season_ppt, year > 2004), 
                          aes(year, rain2d_total))
rainfall_figure +
      geom_boxplot(aes(group = year), notch = T) +
      geom_hline(aes(yintercept = mean(rain2d_total)), linetype = "dashed") +
      theme_bw() +
      xlab("Sample Year") +
      ylab("Total Rainfall (mm)") +
      # ggtitle("Rainfall Estimated at Plots from Rain Gauges (November - May)") +
      theme(panel.grid = element_blank(),
            axis.title = element_text(size = 12),
            axis.text = element_text(size = 10))
# ggsave("figures/rainy-season-ppt.eps", width = 6.25, height = 4.5)
